---
title: "Index page"
author: "Kevin Rue"
date: "`r Sys.Date()`"
output: html_document
---

```{r libraries, include=FALSE}
library(tidyverse)
```

```{r snakemake, include=FALSE}
report_paths <- snakemake@input %>%
  str_remove("reports/")
```

## fishpond

The [fishpond](https://www.bioconductor.org/packages/release/bioc/html/fishpond.html) Bioconductor package was used to import the quantification files produced by [simpleaf](https://simpleaf.readthedocs.io/en/latest/) into R.

A joint report was produced to compare the results of all samples.

```{r, echo=FALSE, results='asis'}
path_all <- grep("fishpond.html", report_paths, value = TRUE)
cat(sprintf("- [All samples](%s)", path_all), sep = "\n")
```

<details>
  <summary>Individual sample reports</summary>
```{r, echo=FALSE, results='asis'}
path_samples <- grep("^fishpond/", report_paths, value = TRUE)
names(path_samples) <- fs::path_ext_remove(basename(path_samples))
cat(sprintf("- [%s](%s)", names(path_samples), path_samples), sep = "\n")
```
</details>

## barcodeRanks

The `barcodeRanks()` function of the [DropletUtils](https://bioconductor.org/packages/release/bioc/html/DropletUtils.html) Bioconductor package was used to dentify the knee and inflection points of the barcode rank curve for each sample.

This section reports on the results of this function, to confirm that reasonable values were obtained.

<details>
  <summary>Individual sample reports</summary>
```{r, echo=FALSE, results='asis'}
path_samples <- grep("^barcodeRanks/", report_paths, value = TRUE)
names(path_samples) <- fs::path_ext_remove(basename(path_samples))
cat(sprintf("- [%s](%s)", names(path_samples), path_samples), sep = "\n")
```
</details>

## emptyDrops

The `emptyDrops()` function of the [DropletUtils](https://bioconductor.org/packages/release/bioc/html/DropletUtils.html) Bioconductor package was used to predict empty droplets for each sample.

This section reports on the results of this function, to guide the decision of droplets to exclude from downstream analysis.

A joint report was produced to compare the results of all samples.

```{r, echo=FALSE, results='asis'}
path_all <- grep("emptyDrops.html", report_paths, value = TRUE)
cat(sprintf("- [All samples](%s)", path_all), sep = "\n")
```

<details>
  <summary>Individual sample reports</summary>
```{r, echo=FALSE, results='asis'}
path_samples <- grep("^emptyDrops/", report_paths, value = TRUE)
names(path_samples) <- fs::path_ext_remove(basename(path_samples))
cat(sprintf("- [%s](%s)", names(path_samples), path_samples), sep = "\n")
```
</details>

## After emptyDrops

This section describes the barcodes remaining after filtering out empty droplets, but before removing droplets with high mitochondrial content.

<details>
  <summary>Individual sample reports</summary>
```{r, echo=FALSE, results='asis'}
path_samples <- grep("^after_emptyDrops/", report_paths, value = TRUE)
names(path_samples) <- fs::path_ext_remove(basename(path_samples))
cat(sprintf("- [%s](%s)", names(path_samples), path_samples), sep = "\n")
```
</details>

## Before scDblFinder

This section describes the barcodes remaining after removing droplets with high mitochondrial content, but before running scDblFinder.

In particular, `scDblFinder()` works best when clustering information is available, to produce artificial doublets made of cells from different clusters.
Consequently, these reports explore the parameter space of clustering algorithms to guide the decision of which clustering results to use as input for `scDblFinder()`.

<details>
  <summary>Individual sample reports</summary>
```{r, echo=FALSE, results='asis'}
path_samples <- grep("^before_scdblfinder/", report_paths, value = TRUE)
names(path_samples) <- fs::path_ext_remove(basename(path_samples))
cat(sprintf("- [%s](%s)", names(path_samples), path_samples), sep = "\n")
```
</details>

## scDblFinder

The `scDblFinder()` function of the [scDblFinder](https://bioconductor.org/packages/release/bioc/html/scDblFinder.html) Bioconductor package was used to predict doublets in each sample.

This section reports on the results of this function, to guide the decision of droplets to exclude from downstream analysis.

**NOTE:** At present, the results are not satisfactory, and none of the predicted doublets were excluded from downstream analysis.

<details>
  <summary>Individual sample reports</summary>
```{r, echo=FALSE, results='asis'}
path_samples <- grep("^scdblfinder/", report_paths, value = TRUE)
names(path_samples) <- fs::path_ext_remove(basename(path_samples))
cat(sprintf("- [%s](%s)", names(path_samples), path_samples), sep = "\n")
```
</details>

## DoubletFinder

The `DoubletFinder()` function of the [DoubletFinder](https://rdrr.io/github/chris-mcginnis-ucsf/DoubletFinder/) R package was used to predict doublets in each sample.

This section reports on the results of this function, to guide the decision of droplets to exclude from downstream analysis.

**NOTE:** At present, the results are not satisfactory, and none of the predicted doublets were excluded from downstream analysis.

<details>
  <summary>Individual sample reports</summary>
```{r, echo=FALSE, results='asis'}
path_samples <- grep("^doubletfinder/", report_paths, value = TRUE)
names(path_samples) <- fs::path_ext_remove(basename(path_samples))
cat(sprintf("- [%s](%s)", names(path_samples), path_samples), sep = "\n")
```
</details>

## Without integration

This section reports on a standard analysis of the data merged without any integration of the samples.

```{r, echo=FALSE, results='asis'}
path_all <- grep("umap.html", report_paths, value = TRUE)
cat(sprintf("- [UMAP without integration](%s)", path_all), sep = "\n")
```

## Integration

This section reports on the integration of the samples using the `fastMNN()` function of the [batchelor](https://bioconductor.org/packages/release/bioc/html/batchelor.html) Bioconductor package.

```{r, echo=FALSE, results='asis'}
path_all <- grep("^hvgs.html", report_paths, value = TRUE)
cat(sprintf("- [Highly variable genes used for PCA and integration](%s)", path_all), sep = "\n")
path_all <- grep("^fastmnn.html", report_paths, value = TRUE)
cat(sprintf("- [fastMNN](%s)", path_all), sep = "\n")
path_all <- grep("^umap_after_integration.html", report_paths, value = TRUE)
cat(sprintf("- [UMAP after integration](%s)", path_all), sep = "\n")
path_all <- grep("^clustering_report.html", report_paths, value = TRUE)
cat(sprintf("- [Clustering after integration](%s)", path_all), sep = "\n")
path_all <- grep("^markers_report.html", report_paths, value = TRUE)
cat(sprintf("- [Cluster markers after integration](%s)", path_all), sep = "\n")
path_all <- grep("^custom_markers_report.html", report_paths, value = TRUE)
cat(sprintf("- [Custom markers on UMAP after integration](%s)", path_all), sep = "\n")
```

## Software environment

```{r, echo=FALSE, results='asis'}
path_all <- grep("packages_r.html", report_paths, value = TRUE)
cat(sprintf("- [R packages](%s)", path_all), sep = "\n")
```
