---
title: "After emptyDrops (sample)"
output: html_document
author: "Kevin Rue"
date: "`r Sys.Date()`"
---

```{r libraries, include=FALSE}
library(BiocParallel)
library(bluster)
library(scales)
library(scater)
library(scran)
library(SingleCellExperiment)
library(tidyverse)
```

```{r setup, include=FALSE}
umi_breaks <- 10^(-1:10)
umi_minor_breaks <- rep(c(1, 2, 5), 12) * (10^rep(-1:10, each = 3))
barcode_breaks <- 10^(-1:10)
barcode_minor_breaks <- 10^rep(-1:10)
```

```{r snakemake, include=FALSE}
sce_rds <- snakemake@input[["sce"]]
mt_tsv <- snakemake@input[["mt"]]
# sce_rds <- "../results/emptyDrops/sce/WPPm048hrs_rep1.rds"
```

## Run parameters

```{r parameters, echo=FALSE}
format_bold <- function(x) sprintf("**%s**", x)
sample_name <- fs::path_ext_remove(basename(sce_rds))
df <- matrix(
  data = c(
    format_bold("Sample name"), sample_name,
    format_bold("Input file"), sce_rds
  ),
  ncol = 2, byrow = TRUE
)
knitr::kable(x = df)
```

## Load SCE object

```{r load_fry, echo=FALSE}
sce <- readRDS(sce_rds)
sce
```

## Load mitochondrial gene list

```{r, echo=FALSE}
mito_gene_ids <- readr::read_tsv(mt_tsv, show_col_types = FALSE)[["gene_id"]]
str(mito_gene_ids)
```

## Barcode statistics

```{r compute_statistics, include=FALSE}
barcode_stats <- tibble(
  umi_sum = colSums(assay(sce, "counts")),
  genes = colSums(assay(sce, "counts") > 0),
  mt_pct = colSums(assay(sce, "counts")[mito_gene_ids, ]) / umi_sum
) %>% 
  filter(umi_sum > 0) %>% 
  arrange(desc(umi_sum)) %>% 
  mutate(rank_umi = row_number()) %>% 
  arrange(desc(genes)) %>% 
  mutate(rank_genes = row_number())
barcode_stats
```

### Range of UMI per barcode

```{r range_umi_sum, echo=FALSE}
range(barcode_stats$umi_sum)
```

### Range of mitochondrial contents

```{r range_mt_pct, echo=FALSE}
range(barcode_stats$mt_pct)
```

### Barcode-Rank Plot

- <span style="color:black">Black</span>: UMI count maximum.
- <span style="color:red">Red</span>: Expected number of cells & UMI count.

```{r barcode_rank_plot, echo=FALSE}
top_stats <- barcode_stats %>%
  slice_max(umi_sum, n = 1, with_ties = FALSE)
bottom_stats <- barcode_stats %>%
  slice_min(umi_sum, n = 1, with_ties = FALSE)
ggplot() +
  geom_line(
    mapping = aes(
      x = rank_umi,
      y = umi_sum
    ),
    data = barcode_stats) +
  geom_label(
    aes(
      x = rank_umi,
      y = umi_sum,
      label = format(umi_sum, big.mark = ",")
    ),
    data = top_stats,
    alpha = 0.5,
    hjust = 0
  ) +
  geom_label(
    aes(
      x = rank_umi,
      y = umi_sum,
      label = format(umi_sum, big.mark = ",")
    ),
    data = bottom_stats,
    alpha = 0.5,
    hjust = 1
  ) +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_log10(
    limits = c(1, NA),
    breaks = umi_breaks,
    minor_breaks = umi_minor_breaks,
    labels = label_number(scale_cut = cut_short_scale())
  ) +
  theme_bw() +
  labs(
    x = "Barcodes",
    y = "UMI Counts",
    title = paste0("Sample (", sample_name, ")"),
    subtitle = paste0("Barcodes (", format(ncol(sce), big.mark = ","), ")")
  )
```

### Features-Rank Plot

- <span style="color:black">Black</span>: Detected genes maximum.
- <span style="color:red">Red</span>: Expected number of cells & detected genes.

```{r feature_rank_plot, echo=FALSE}
top_stats <- barcode_stats %>%
  slice_max(genes, n = 1, with_ties = FALSE)
bottom_stats <- barcode_stats %>%
  slice_min(genes, n = 1, with_ties = FALSE)
ggplot() +
  geom_line(aes(x = rank_genes, y = genes), barcode_stats) +
  geom_label(
    aes(
      x = rank_genes,
      y = genes,
      label = format(genes, big.mark = ",")
    ),
    data = top_stats,
    alpha = 0.5,
    hjust = 0
  ) +
  geom_label(
    aes(
      x = rank_genes,
      y = genes,
      label = format(genes, big.mark = ",")
    ),
    data = bottom_stats,
    alpha = 0.5,
    hjust = 1
  ) +
  scale_x_log10(
    breaks = barcode_breaks,
    minor_breaks = barcode_minor_breaks,
    labels = label_number(scale_cut = cut_short_scale())
  ) +
  scale_y_continuous(
    limits = c(0, NA),
    labels = label_number(scale_cut = cut_short_scale())
  ) +
  theme_bw() +
  labs(
    x = "Barcodes",
    y = "Genes detected",
    title = paste0("Sample (", sample_name, ")"),
    subtitle = paste0("Barcodes (", format(ncol(sce), big.mark = ","), ")")
  )
```

## Mitochondrial content

```{r mito_histogram, echo=FALSE}
barcode_stats %>% 
  ggplot(aes(x = mt_pct)) +
  geom_histogram(binwidth = 0.01, colour = "black", fill = "grey", linewidth = 0.1) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, NA)) +
  scale_x_continuous(labels = label_percent()) +
  scale_y_continuous(labels = label_comma()) +
  theme_bw() +
  labs(
    y = "Barcodes",
    x = "Proportion of mitochondrial counts",
    title = paste0("Sample (", sample_name, ")"),
    subtitle = paste0("Barcodes (", format(ncol(sce), big.mark = ","), ")")
  )
```

## Log-normalise

Necessary prior to HVGs.

```{r, echo=FALSE}
sce <- logNormCounts(
  x = sce,
  BPPARAM = MulticoreParam(workers = snakemake@threads)
)
sce
```

## HVGs

Recommended for PCA.

```{r, include=FALSE}
dec <- modelGeneVar(
  x = sce,
  BPPARAM = MulticoreParam(workers = snakemake@threads)
)
dec
```

```{r, include=FALSE}
hvgs <- getTopHVGs(dec, n = 500)
length(hvgs)
```

```{r, echo=FALSE}
dec[["selected"]] <- factor(rownames(dec) %in% hvgs, c(TRUE, FALSE))
ggplot(dec) +
  geom_point(aes(mean, total, colour = selected)) +
  geom_point(aes(mean, tech), color = "black") +
  scale_colour_manual(values = c("TRUE" = "deepskyblue", "FALSE" = "coral2")) +
  theme_bw()
```

## PCA

```{r, include=FALSE}
sce <- fixedPCA(
  x = sce,
  subset.row = hvgs,
  BPPARAM = MulticoreParam(workers = snakemake@threads)
)
sce
```

```{r, echo=FALSE}
plot_data <- tibble(
  percentVar = attr(reducedDim(sce, "PCA"), "percentVar")
) %>% 
  mutate(
    PC = factor(row_number())
  )
ggplot(plot_data, aes(PC, percentVar)) +
  geom_col(color = "black", fill = "grey", linewidth = 0.1) +
  scale_y_continuous(limits = c(0, NA)) +
  theme_bw()
```

## UMAP

```{r, include=FALSE}
sce <- runUMAP(
  x = sce,
  dimred = "PCA",
  n_dimred = 20,
  BPPARAM = MulticoreParam(workers = snakemake@threads)
)
sce
```

```{r, echo=FALSE, fig.height=6, fig.width=8}
reducedDim(sce, "UMAP") %>% 
  as_tibble() %>% 
  ggplot(aes(UMAP1, UMAP2)) +
  geom_point(size = 0.01) +
  theme_bw()
```

## Clustering

```{r, echo=FALSE}
colData(sce)[["cluster"]] <- clusterRows(
  x = reducedDim(sce, "UMAP"),
  BLUSPARAM = DbscanParam(
    min.pts = 500,
    core.prop = 0.95
  )
)
table(colData(sce)[["cluster"]])
```

```{r, echo=FALSE}
reducedDim(sce, "UMAP") %>% 
  as_tibble() %>% 
  bind_cols(colData(sce) %>% as_tibble()) %>% 
  ggplot(aes(UMAP1, UMAP2, colour = cluster)) +
  geom_point(size = 0.01) +
  theme_bw()
```

## Session info

`r date()`

Bioconductor version `r BiocManager::version()`

```{r session_info, echo=FALSE}
sessionInfo()
```
