---
title: "Fishpond (merged)"
output: html_document
author: "Kevin Rue"
date: "`r Sys.Date()`"
---

```{r libraries, include=FALSE}
library(cowplot)
library(ggplot2)
library(ggrastr)
library(scales)
library(SingleCellExperiment)
library(stringr)
library(tidyverse)
```

```{r constants, include=FALSE}
umi_breaks <- 10^(-1:10)
umi_minor_breaks <- rep(c(1, 2, 5), 12) * (10^rep(-1:10, each = 3))
barcode_breaks <- 10^(-1:10)
barcode_minor_breaks <- 10^rep(-1:10)
```

```{r snakemake, include=FALSE}
rds_file <- snakemake@input[["rds"]]
mt_tsv <- snakemake@input[["mt"]]
```

## Run parameters

```{r parameters, echo=FALSE}
format_bold <- function(x) sprintf("**%s**", x)
df <- matrix(
  data = c(
    format_bold("Input file"), rds_file
  ),
  ncol = 2, byrow = TRUE
)
knitr::kable(x = df)
```

## Load SingleCellExperiment object

```{r load_sce, echo=FALSE}
sce <- readRDS(rds_file)
sce
```

## Load mitochondrial gene list

```{r mito_gene_ids, echo=FALSE ,results='asis'}
mito_gene_ids <- read_tsv(mt_tsv, show_col_types = FALSE)[["gene_id"]]
cat("Mitochondrial genes loaded: ", length(mito_gene_ids), "\n")
```

## Barcode statistics

```{r barcode_metadata, include=FALSE}
timepoint_levels <- c("m048hrs", "m024hrs", "p000hrs", "p024hrs", "p048hrs", "p072hrs", "p096hrs", "p120hrs")
replicate_levels <- c("1", "2")
barcode_stats <- tibble(
  barcode = colnames(sce),
  sample = colData(sce)[["sample"]]
) %>% 
  bind_cols(
    str_match(string = colData(sce)[["sample"]], pattern = "WPP(?<timepoint>[mp][[:digit:]]{3}hrs)_rep(?<replicate>[[:digit:]]{1})")[, -1]
  ) %>% 
  mutate(
    timepoint = factor(timepoint, timepoint_levels),
    replicate = factor(replicate, replicate_levels)
  )
```

```{r compute_statistics, echo=FALSE}
barcode_stats <- barcode_stats %>% 
  mutate(
    umi_sum = colSums(assay(sce, "counts")),
    genes = colSums(assay(sce, "counts") > 0),
    mt_pct = colSums(assay(sce, "counts")[mito_gene_ids, ]) / umi_sum
  )
barcode_stats <- barcode_stats %>%
  arrange(desc(umi_sum)) %>%
  group_by(sample) %>%
  mutate(rank_umi = row_number()) %>%
  ungroup()
barcode_stats <- barcode_stats %>%
  arrange(desc(genes)) %>%
  group_by(sample) %>%
  mutate(rank_genes = row_number()) %>%
  ungroup()
barcode_stats
rm(sce)
```

- **barcode:** Barcode
- **sample:** Sample name
- **timepoint:** Developmental time point
- **replicate:** Experimental replicate
- **umi_sum:** Total UMIs detected
- **genes:** Total genes detected
- **rank_umi:** Ranking by decreasing UMI detected
- **rank_genes:** Ranking by decreasing genes detected
- **mt_pct:** Proportion of UMIs mapping to mitochondrial genes

## Plots

### Total UMI

```{r umi_rank_facet_sample, fig.width=9, fig.height=7, echo=FALSE}
ggplot(barcode_stats, aes(x = rank_umi, y = umi_sum)) +
  geom_line() +
  facet_wrap(~sample) +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_log10(breaks = umi_breaks, minor_breaks = umi_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  labs(
    y = "UMI Counts",
    x = "Barcodes"
  ) +
  theme_bw()
```

```{r umi_rank_colour_sample, fig.width=12, fig.height=5, echo=FALSE}
gg1 <- ggplot(barcode_stats, aes(x = rank_umi, y = umi_sum, colour = sample)) +
  geom_line() +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_log10(breaks = umi_breaks, minor_breaks = umi_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  guides(
    colour = guide_legend(override.aes = list(linewidth = 3))
  ) +
  labs(
    y = "UMI Counts",
    x = "Barcodes"
  ) +
  theme_bw()
gg2 <- ggplot(barcode_stats, aes(x = rank_umi, y = umi_sum, colour = sample)) +
  geom_line() +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_log10(breaks = umi_breaks, minor_breaks = umi_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  coord_cartesian(ylim = c(1, 10E3), xlim = c(10E3, NA)) +
  guides(
    colour = guide_legend(override.aes = list(linewidth = 3))
  ) +
  labs(
    y = "UMI Counts",
    x = "Barcodes"
  ) +
  theme_bw()
plot_grid(gg1, gg2)
rm(gg1, gg2)
```

### Genes detected

```{r genes_rank_facet_sample, fig.width=9, fig.height=7, echo=FALSE}
ggplot(barcode_stats, aes(x = rank_genes, y = genes)) +
  geom_line() +
  facet_wrap(~sample) +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(labels = label_comma()) +
  guides(
    colour = guide_legend(override.aes = list(linewidth = 3))
  ) +
  labs(
    y = "Genes",
    x = "Barcodes"
  ) +
  theme_bw()
```

```{r genes_rank_colour_sample, fig.width=12, fig.height=5, echo=FALSE}
gg1 <- ggplot(barcode_stats, aes(x = rank_genes, y = genes, colour = sample)) +
  geom_line() +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(labels = label_comma()) +
  guides(
    colour = guide_legend(override.aes = list(linewidth = 3))
  ) +
  labs(
    y = "Genes",
    x = "Barcodes"
  ) +
  theme_bw()
gg2 <- ggplot(barcode_stats, aes(x = rank_genes, y = genes, colour = sample)) +
  geom_line() +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(labels = label_comma()) +
  coord_cartesian(ylim = c(0, 2E3), xlim = c(10E3, NA)) +
  guides(
    colour = guide_legend(override.aes = list(linewidth = 3))
  ) +
  labs(
    y = "Genes",
    x = "Barcodes"
  ) +
  theme_bw()
plot_grid(gg1, gg2)
rm(gg1, gg2)
```

### Mitochondrial content

Note: Filtering droplets with more than 5,000 UMI counts and 1,000 genes detected.

```{r mt_density, echo=FALSE, fig.height=5, fig.width=9}
barcode_stats %>%
  filter(umi_sum > 5E3 & genes > 1E3) %>%
  ggplot() +
  geom_density(aes(x = mt_pct, colour = sample)) +
  coord_cartesian(xlim = c(0, 1)) +
  scale_x_continuous(labels = label_percent()) +
  # scale_y_continuous(labels = label_comma()) +
  theme_bw() +
  labs(
    y = "Barcodes",
    x = "Proportion of mitochondrial counts"
  )
```

```{r mt_scatter_umi, echo=FALSE, fig.height=5, fig.width=8}
barcode_stats %>%
  filter(umi_sum > 5E3 & genes > 1E3) %>%
  ggplot() +
  rasterise(
    geom_point(
      aes(x = umi_sum, y = mt_pct),
      size = 0.001
    ),
    dpi = 300
  ) +
  theme_bw() +
  labs(
    y = "MT %",
    x = "UMI"
  )
```

To do:

- overlaid density plot of mitochondrial content
- scatter plot of mitochondrial content vs. UMI counts

### Total UMI and genes detected

```{r umi_genes_scatterplot_colour_sample, fig.width=9, fig.height=6, echo=FALSE}
barcode_stats %>%
  filter(umi_sum < 20E3 & genes < 5E3) %>%
  ggplot() +
  facet_wrap(~sample) +
  rasterise(
    geom_point(
      mapping = aes(x = umi_sum, y = genes),
      size = 0.1
    ),
    dpi = 300
  ) +
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  theme_bw() +
  labs(
    x = "UMI counts",
    y = "Genes detected"
  )
```

## Session info

`r date()`

Bioconductor version `r BiocManager::version()`

```{r session_info, echo=FALSE}
sessionInfo()
```
