---
title: "rox1:rox2 coexpression"
author: "Kevin Rue"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(SummarizedExperiment)
library(tidyverse)
```


```{r}
sce <- readRDS("results/before_scdblfinder/WPPp120hrs_rep1.rds")
sce
```

```{r}
rox1_name <- "lncRNA:roX1"
rox2_name <- "lncRNA:roX2"
```

```{r}
rox1_id <- rowData(sce) %>% 
  as_tibble() %>% 
  filter(gene_name == rox1_name) %>% 
  pull(gene_id)
rox2_id <- rowData(sce) %>% 
  as_tibble() %>% 
  filter(gene_name == rox2_name) %>% 
  pull(gene_id)
```

First, look at log-normalised counts, as those are meant to, well, normalise, between cells.

```{r scatter_logcounts}
gg_data <- tibble(
  rox1 = assay(sce, "logcounts")[rox1_id, ],
  rox2 = assay(sce, "logcounts")[rox2_id, ],
)
ggplot() +
  geom_point(
    mapping = aes(x = rox1, y = rox2),
    data = gg_data
  ) +
  theme_bw()
```

Then, look at raw counts, as we are comparing the two genes within each cell, so:

* normalisation does not change the relative expression of the two genes within each cell
* we are not comparing between cell
* we get a clearer picture of the actual value observed for each gene in each cell

```{r scatter_counts, warning=FALSE}
gg_data <- tibble(
  rox1 = assay(sce, "counts")[rox1_id, ],
  rox2 = assay(sce, "counts")[rox2_id, ],
)
ggplot() +
  geom_point(
    mapping = aes(x = rox1, y = rox2),
    data = gg_data
  ) +
  scale_x_log10() + scale_y_log10() +
  theme_bw()
```

