---
title: "Quality control of barcodes"
author: "Kevin Rue"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
umi_breaks <- 10^(-1:10)
umi_minor_breaks <- rep(c(1, 2, 5), 12) * (10^rep(-1:10, each = 3))
barcode_breaks <- 10^(-1:10)
barcode_minor_breaks <- 10^rep(-1:10)
dir.create("results/alevin-reports/data/", showWarnings = FALSE)
```

## Libraries

```{r, message=FALSE}
library(cowplot)
library(ggplot2)
library(ggrastr)
library(scales)
library(SingleCellExperiment)
library(stringr)
library(tidyverse)
```

## Load sce

```{r load_sce, echo=FALSE}
sce <- readRDS(snakemake@input[["rds"]])
sce
```

## Barcode statistics

```{r barcode_metadata, include=FALSE}
timepoint_levels <- c("m048hrs", "m024hrs", "p000hrs", "p024hrs", "p048hrs", "p072hrs", "p096hrs", "p120hrs")
replicate_levels <- c("1", "2")
barcode_stats <- tibble(
  barcode = colnames(sce),
  sample = colData(sce)[["sample"]]
) %>% 
  bind_cols(
    str_match(string = colData(sce)[["sample"]], pattern = "WPP(?<timepoint>[mp][[:digit:]]{3}hrs)_rep(?<replicate>[[:digit:]]{1})")[, -1]
  ) %>% 
  mutate(
    timepoint = factor(timepoint, timepoint_levels),
    replicate = factor(replicate, replicate_levels)
  )
```

- barcode: Barcode
- sample: Sample name
- timepoint: Developmental time point
- replicate: Experimental replicate
- umi_sum: Total UMIs detected
- genes: Total genes detected
- rank_umi: Ranking by decreasing UMI detected
- rank_genes: Ranking by decreasing genes detected

```{r compute_statistics, include=FALSE}
barcode_stats$umi_sum <- colSums(assay(sce, "counts"))
barcode_stats$genes <- colSums(assay(sce, "counts") > 0)
barcode_stats <- barcode_stats %>% 
  arrange(desc(umi_sum)) %>% 
  group_by(sample) %>% 
  mutate(rank_umi = row_number()) %>% 
  ungroup()
barcode_stats <- barcode_stats %>% 
  arrange(desc(genes)) %>% 
  group_by(sample) %>% 
  mutate(rank_genes = row_number()) %>% 
  ungroup()
barcode_stats
rm(sce)
```

```{r write_barcode_stats, include=FALSE}
write_tsv(barcode_stats, "results/alevin-reports/data/barcode_stats.tsv.gz")
```

## Proposed QC cutoffs

```{r set_cutoffs}
umi_cutoff_low <- snakemake@params[["umi_cutoff_low"]]
genes_cutoff_low <- snakemake@params[["genes_cutoff_low"]]
umi_cutoff_final <- snakemake@params[["umi_cutoff_final"]]
genes_cutoff_final <- snakemake@params[["genes_cutoff_final"]]
```

## Plots

### Total UMI

```{r umi_rank_facet_sample, fig.width=9, fig.height=7, echo=FALSE}
filtered_stats_final <- barcode_stats %>% 
  group_by(sample) %>% 
  summarise(barcodes = sum(umi_sum >= umi_cutoff_final))
ggplot(barcode_stats, aes(x = rank_umi, y = umi_sum)) +
  geom_line() +
  geom_hline(yintercept = umi_cutoff_low, linetype = "dashed", linewidth = 0.5, colour = "orange") +
  geom_hline(yintercept = umi_cutoff_final, linetype = "dashed", linewidth = 0.5, colour = "red") +
  geom_label(aes(x = 10, y = umi_cutoff_final, label = format(barcodes, big.mark = ",")), filtered_stats_final, colour = "red") +
  facet_wrap(~sample) +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_log10(breaks = umi_breaks, minor_breaks = umi_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  theme_bw() +
  labs(
    y = "UMI Counts",
    x = "Barcodes"
  )
```

```{r umi_rank_colour_sample, fig.width=12, fig.height=5, echo=FALSE}
gg1 <- ggplot(barcode_stats, aes(x = rank_umi, y = umi_sum, colour = sample)) +
  geom_line() +
  geom_hline(yintercept = umi_cutoff_low, linetype = "dashed", linewidth = 0.5, colour = "orange") +
  geom_hline(yintercept = umi_cutoff_final, linetype = "dashed", linewidth = 0.5, colour = "red") +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_log10(breaks = umi_breaks, minor_breaks = umi_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  theme_bw() +
  labs(
    y = "UMI Counts",
    x = "Barcodes"
  )
gg2 <- ggplot(barcode_stats, aes(x = rank_umi, y = umi_sum, colour = sample)) +
  geom_line() +
  geom_hline(yintercept = umi_cutoff_low, linetype = "dashed", linewidth = 0.5, colour = "orange") +
  geom_hline(yintercept = umi_cutoff_final, linetype = "dashed", linewidth = 0.5, colour = "red") +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_log10(breaks = umi_breaks, minor_breaks = umi_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  coord_cartesian(ylim = c(1, 10E3), xlim = c(10E3, NA)) +
  theme_bw() +
  labs(
    y = "UMI Counts",
    x = "Barcodes"
  )
plot_grid(gg1, gg2)
rm(gg1, gg2)
```

```{r umi_histogram_facet_sample, fig.width=9, fig.height=6, echo=FALSE}
ggplot(barcode_stats %>% filter(umi_sum < 20E3), aes(x = umi_sum)) +
  facet_wrap(~sample) +
  geom_histogram(color = "black", fill = "white", binwidth = 100, position = "identity", alpha = 0.5, linewidth = 0.1) +
  geom_vline(xintercept = umi_cutoff_low, linetype = "dashed", linewidth = 0.5, colour = "orange") +
  geom_vline(xintercept = umi_cutoff_final, linetype = "dashed", linewidth = 0.5, colour = "red") +
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_log10(labels = label_number(scale_cut = cut_short_scale())) +
  theme_bw() +
  labs(
    x = "Barcodes",
    y = "UMI counts"
  )
```

### Genes detected

```{r genes_rank_facet_sample, fig.width=9, fig.height=7, echo=FALSE}
filtered_stats_final <- barcode_stats %>% 
  group_by(sample) %>% 
  summarise(barcodes = sum(genes >= genes_cutoff_final))
ggplot(barcode_stats, aes(x = rank_genes, y = genes)) +
  geom_line() +
  geom_hline(yintercept = genes_cutoff_low, linetype = "dashed", linewidth = 0.5, colour = "orange") +
  geom_hline(yintercept = genes_cutoff_final, linetype = "dashed", linewidth = 0.5, colour = "red") +
  geom_label(aes(x = 10, y = genes_cutoff_final, label = format(barcodes, big.mark = ",")), filtered_stats_final, colour = "red") +
  facet_wrap(~sample) +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(labels = label_comma()) +
  theme_bw() +
  labs(
    y = "Genes",
    x = "Barcodes"
  )
```

```{r genes_rank_colour_sample, fig.width=12, fig.height=5, echo=FALSE}
gg1 <- ggplot(barcode_stats, aes(x = rank_genes, y = genes, colour = sample)) +
  geom_line() +
  geom_hline(yintercept = genes_cutoff_low, linetype = "dashed", linewidth = 0.5, colour = "orange") +
  geom_hline(yintercept = genes_cutoff_final, linetype = "dashed", linewidth = 0.5, colour = "red") +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(labels = label_comma()) +
  theme_bw() +
  labs(
    y = "Genes",
    x = "Barcodes"
  )
gg2 <- ggplot(barcode_stats, aes(x = rank_genes, y = genes, colour = sample)) +
  geom_line() +
  geom_hline(yintercept = genes_cutoff_low, linetype = "dashed", linewidth = 0.5, colour = "orange") +
  geom_hline(yintercept = genes_cutoff_final, linetype = "dashed", linewidth = 0.5, colour = "red") +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(labels = label_comma()) +
  coord_cartesian(ylim = c(0, 2E3), xlim = c(10E3, NA)) +
  theme_bw() +
  labs(
    y = "Genes",
    x = "Barcodes"
  )
plot_grid(gg1, gg2)
rm(gg1, gg2)
```

```{r genes_histogram_colour_sample, fig.width=9, fig.height=6, echo=FALSE}
ggplot(barcode_stats %>% filter(genes < 5E3), aes(x = genes)) +
  facet_wrap(~sample) +
  geom_histogram(color = "black", fill = "white", binwidth = 25, position = "identity", alpha = 0.5, linewidth = 0.1) +
  scale_y_log10(breaks = umi_breaks, minor_breaks = umi_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  geom_vline(xintercept = genes_cutoff_low, linetype = "dashed", linewidth = 0.5, colour = "orange") +
  geom_vline(xintercept = genes_cutoff_final, linetype = "dashed", linewidth = 0.5, colour = "red") +
  theme_bw() +
  labs(
    x = "Genes",
    y = "Barcodes"
  )
```

### Total UMI and genes detected

```{r umi_genes_scatterplot_colour_sample, fig.width=9, fig.height=6, echo=FALSE}
ggplot(barcode_stats %>% filter(umi_sum < 20E3 & genes < 5E3), aes(x = genes, y = umi_sum)) +
  facet_wrap(~sample) +
  geom_point_rast(size = 0.1) +
  geom_vline(xintercept = genes_cutoff_low, linetype = "dashed", linewidth = 0.5, colour = "orange") +
  geom_vline(xintercept = genes_cutoff_final, linetype = "dashed", linewidth = 0.5, colour = "red") +
  geom_hline(yintercept = umi_cutoff_low, linetype = "dashed", linewidth = 0.5, colour = "orange") +
  geom_hline(yintercept = umi_cutoff_final, linetype = "dashed", linewidth = 0.5, colour = "red") +
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  theme_bw() +
  labs(
    x = "UMI counts",
    y = "Barcodes"
  )
```

## Apply filters

### Permissive filters

```{r filters_permissive}
barcode_stats %>% 
  group_by(sample) %>% 
  summarise(
    umi_filter = sum(umi_sum >= umi_cutoff_low),
    genes_filter = sum(genes >= genes_cutoff_low),
    all_filters = sum(umi_sum >= umi_cutoff_low & genes >= genes_cutoff_low)
  ) %>%
  knitr::kable(format.args = list(big.mark = ","))
```

Total: `r format(with(barcode_stats, sum(umi_sum >= umi_cutoff_low & genes >= genes_cutoff_low)), big.mark = ",")`

### Final filters

```{r filters_final}
barcode_stats %>% 
  group_by(sample) %>% 
  summarise(
    umi_filter = sum(umi_sum >= umi_cutoff_final),
    genes_filter = sum(genes >= genes_cutoff_final),
    all_filters = sum(umi_sum >= umi_cutoff_final & genes >= genes_cutoff_final)
  ) %>%
  knitr::kable(format.args = list(big.mark = ","))
```

Total: `r format(with(barcode_stats, sum(umi_sum >= umi_cutoff_final & genes >= genes_cutoff_final)), big.mark = ",")`
