---
title: "Individual sample quality control"
output: html_document
author: "Kevin Rue"
date: "`r Sys.Date()`"
---

## Sample information


```{r libraries, include=FALSE}
library(cowplot)
library(ggplot2)
library(scales)
library(SingleCellExperiment)
library(tidyverse)
umi_breaks <- 10^(-1:10)
umi_minor_breaks <- rep(c(1, 2, 5), 12) * (10^rep(-1:10, each = 3))
barcode_breaks <- 10^(-1:10)
barcode_minor_breaks <- 10^rep(-1:10)
```

```{r parameters, include=FALSE}
expect_cells <- snakemake@params[["expect_cells"]]
```

- Name: `r basename(snakemake@input[["sce"]])`
- Expected number of cells: `r format(expect_cells, big.mark = ",")`

## Load SCE object

```{r load_fry, echo=FALSE}
#| message: false
sce <- readRDS(snakemake@input[["sce"]])
sce
```

## Load mitochondrial gene list

```{r, echo=FALSE}
mito_gene_ids <- readr::read_tsv("config/mitochdondrial_genes.tsv", show_col_types = FALSE)[["gene_id"]]
str(mito_gene_ids)
```

## Barcode statistics

```{r compute_statistics, include=FALSE}
barcode_stats <- tibble(
  umi_sum = colSums(assay(sce, "counts")),
  genes = colSums(assay(sce, "counts") > 0),
  mt_pct = colSums(assay(sce, "counts")[mito_gene_ids, ]) / umi_sum
) %>% 
  filter(umi_sum > 0) %>% 
  arrange(desc(umi_sum)) %>% 
  mutate(rank_umi = row_number()) %>% 
  arrange(desc(genes)) %>% 
  mutate(rank_genes = row_number())
barcode_stats
rm(sce)
```

### Range of UMI per barcode

```{r range_umi_sum, echo=FALSE}
range(barcode_stats$umi_sum)
```

### Range of mitochondrial contents

```{r range_mt_pct, echo=FALSE}
range(barcode_stats$mt_pct)
```

### Barcode-Rank Plot

Labels:

- Maximum UMI count (black).
- Minimum UMI counts for expected number of cells (red).
- Expected number of cells (red).

```{r barcode_rank_plot, echo=FALSE}
expect_cells_min_umi <- barcode_stats %>% 
	arrange(desc(umi_sum)) %>% 
	slice_head(n = expect_cells) %>% 
	slice_tail(n = 1) %>% 
	pull(umi_sum)
top_stats <- barcode_stats %>%
  slice_max(umi_sum, n = 1)
ggplot() +
  geom_line(aes(x = rank_umi, y = umi_sum), barcode_stats) +
  geom_label(aes(x = rank_umi, y = umi_sum, label = format(umi_sum, big.mark = ",")), data = top_stats, alpha = 0.5, hjust = 0) +
  geom_hline(yintercept = expect_cells_min_umi, linetype = "dashed", linewidth = 0.5, colour = "red") +
  geom_vline(xintercept = expect_cells, linetype = "dashed", linewidth = 0.5, colour = "red") +
  geom_label(aes(x = 1, y = expect_cells_min_umi, label = format(expect_cells_min_umi, big.mark = ",")), data = tibble(), colour = "red", hjust = 0) +
	geom_label(aes(x = expect_cells, y = 1, label = format(expect_cells, big.mark = ",")), data = tibble(), colour = "red", vjust = 0) +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_log10(breaks = umi_breaks, minor_breaks = umi_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  theme_bw() +
  labs(
    x = "Barcodes",
    y = "UMI Counts"
  )
```

### Features-Rank Plot

Labels:

- Maximum genes detected (black).
- Minimum genes detected for expected number of cells (red).
- Expected number of cells (red).

```{r feature_rank_plot, echo=FALSE}
expect_cells_min_genes <- barcode_stats %>% 
	arrange(desc(genes)) %>% 
	slice_head(n = expect_cells) %>% 
	slice_tail(n = 1) %>% 
	pull(genes)
top_stats <- barcode_stats %>%
  slice_max(genes, n = 1)
ggplot() +
  geom_line(aes(x = rank_genes, y = genes), barcode_stats) +
  geom_label(aes(x = rank_genes, y = genes, label = format(genes, big.mark = ",")), data = top_stats, alpha = 0.5, hjust = 0) +
  geom_hline(yintercept = expect_cells_min_genes, linetype = "dashed", linewidth = 0.5, colour = "red") +
  geom_vline(xintercept = expect_cells, linetype = "dashed", linewidth = 0.5, colour = "red") +
  geom_label(aes(x = 1, y = expect_cells_min_genes, label = format(expect_cells_min_genes, big.mark = ",")), data = tibble(), colour = "red", hjust = 0) +
	geom_label(aes(x = expect_cells, y = 1, label = format(expect_cells, big.mark = ",")), data = tibble(), colour = "red", vjust = 0) +
  scale_x_log10(breaks = barcode_breaks, minor_breaks = barcode_minor_breaks, labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  theme_bw() +
  labs(
    x = "Barcodes",
    y = "Genes detected"
  )
```

## Mitochondrial content

Mitochondrial content is displayed only for barcodes with:

- at least `r format(expect_cells_min_umi, big.mark = ",")` UMI counts
- at least `r format(expect_cells_min_genes, big.mark = ",")` genes detected

```{r mito_histogram, echo=FALSE}
barcode_stats %>% 
  filter(umi_sum >= expect_cells_min_umi) %>% 
  filter(genes >= expect_cells_min_genes) %>%
  ggplot(aes(x = mt_pct)) +
  geom_histogram(binwidth = 0.01, colour = "black", fill = "grey", linewidth = 0.1) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, NA)) +
  scale_x_continuous(labels = label_percent()) +
  scale_y_continuous(labels = label_comma()) +
  theme_bw() +
  labs(
    y = "Barcodes",
    x = "Proportion of mitochondrial counts"
  )
```

## Session info

`r date()`

Bioconductor version `r BiocManager::version()`

```{r session_info, echo=FALSE}
sessionInfo()
```
